---
title: "Shifts in mutation effects among variant backgrounds"
author: "Tyler Starr"
date: "6/9/2022"
output:
  github_document:
    toc: true
    html_preview: false
editor_options: 
  chunk_output_type: inline
---
This notebook analyzes sites whose mutation effects deviate most strongly among the variant RBD backgrounds.

```{r setup, message=FALSE, warning=FALSE, error=FALSE}
require("knitr")
knitr::opts_chunk$set(echo = T)
knitr::opts_chunk$set(dev.args = list(png = list(type = "cairo")))

#list of packages to install/load
packages = c("yaml","data.table","tidyverse","gridExtra","bio3d","ggridges","ggrepel","GGally")
#install any packages not already installed
installed_packages <- packages %in% rownames(installed.packages())
if(any(installed_packages == F)){
  install.packages(packages[!installed_packages])
}
#load packages
invisible(lapply(packages, library, character.only=T))

#read in config file
config <- read_yaml("config.yaml")

#make output directory
if(!file.exists(config$epistatic_shifts_dir)){
  dir.create(file.path(config$epistatic_shifts_dir))
}

#make pdb output directory
if(!file.exists(paste(config$epistatic_shifts_dir,"/pdbs/",sep=""))){
  dir.create(file.path(paste(config$epistatic_shifts_dir,"/pdbs/",sep="")))
}
```
Session info for reproducing environment:
```{r print_sessionInfo}
sessionInfo()
```

Define colorblind-friendly palette
```{r define_color_palette}
# The palette with grey:
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", 
               "#0072B2", "#D55E00", "#CC79A7")
# The palette with black
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", 
                "#0072B2", "#D55E00", "#CC79A7")
```

## Setup

Read in tables of mutant measurements from current study and prior VOCs DMS datasets.

```{r input_data}
dt <- data.table(read.csv(file=config$final_variant_scores_mut_file,stringsAsFactors=F))
```


## Calculate site-wise Jensen-Shannon divergence, a metric of divergence in site-specific mutational profiles

For each pair of backgrounds, at each site I want to compute the Jensen-Shannon divergence between the profile of mutation effects of all mutations at the site. (Remove any measurements determined for <3 or <5 bc to avoid measurements with lower precision driving noise in the global JSD metric.)


```{r setup_table}
#define a minbc bind measurement that requires three barcodes be averaged for a final determination, otherwise change to NA
dt[,bind_min3bc := bind]
dt[n_bc_bind < 3, bind_min3bc := NA]

dt[,bind_min5bc := bind]
dt[n_bc_bind < 5, bind_min5bc := NA]

#define a minbc expr measurement that requires three barcodes be averaged for a final determination, otherwise change to NA
dt[,expr_min3bc := expr]
dt[n_bc_expr < 3, expr_min3bc := NA]

dt[,expr_min5bc := expr]
dt[n_bc_expr < 5, expr_min5bc := NA]

#define a function for computing J-S divergence/distance between two affinity vectors (where affinity is given currently as log10-Kd)
JS <- function(vec1,vec2){
  vec1_pair <- vec1[!is.na(vec1) & !is.na(vec2)]
  vec2_pair <- vec2[!is.na(vec1) & !is.na(vec2)]
  pi1 <- 10^(vec1_pair)/sum(10^(vec1_pair))
  pi2 <- 10^(vec2_pair)/sum(10^(vec2_pair))
  n <- 0.5 * (pi1+pi2)
  JS <- 0.5 * (sum(pi1*log(pi1/n)) + sum(pi2*log(pi2/n)))
  #return(sqrt(JS)) #if doing distance
  return(JS) #if doing divergence
}

#first, for bind measurements
#data table for storing difference in correlation in profiles between bg pairs at each site
#generate table with all combinations of bg_1 and bg_2 for each site
diffs_bind <- data.table(expand.grid(site=unique(dt$position),bg_2=c("Wuhan-Hu-1_v2","Omicron_BA1","Omicron_BA2","Wuhan-Hu-1_v1","Alpha","Beta","Delta","Eta"),bg_1=c("Wuhan-Hu-1_v2","Omicron_BA1","Omicron_BA2","Wuhan-Hu-1_v1","Alpha","Beta","Delta","Eta")))

#remove duplicates where bg_1 and _2 the same
diffs_bind <- diffs_bind[bg_1 != bg_2,]

#loop through and compute JSD for each site for each pair of bgs, for bind metric
diffs_bind$JSD <- as.numeric(NA) #jensen-shannon divergence, from raw bind values (lower limit 5)
diffs_bind$JSD_min3bc <- as.numeric(NA) #jensen-shannon divergence, require a minimum of 3 bcs averaged
diffs_bind$JSD_min5bc <- as.numeric(NA) #jensen-shannon divergence, require a minimum of 5 bcs averaged
for(i in 1:nrow(diffs_bind)){
  x_uncens <- dt[target==diffs_bind[i,bg_1] & position==diffs_bind[i,site],bind]
  y_uncens <- dt[target==diffs_bind[i,bg_2] & position==diffs_bind[i,site],bind]
  x_min3bc <- dt[target==diffs_bind[i,bg_1] & position==diffs_bind[i,site],bind_min3bc]
  y_min3bc <- dt[target==diffs_bind[i,bg_2] & position==diffs_bind[i,site],bind_min3bc]
  x_min5bc <- dt[target==diffs_bind[i,bg_1] & position==diffs_bind[i,site],bind_min5bc]
  y_min5bc <- dt[target==diffs_bind[i,bg_2] & position==diffs_bind[i,site],bind_min5bc]
  diffs_bind[i,JSD := JS(x_uncens,y_uncens)]
  diffs_bind[i,JSD_min3bc := JS(x_min3bc,y_min3bc)]
  diffs_bind[i,JSD_min5bc := JS(x_min3bc,y_min5bc)]
}

#repeat for expr measurements
#data table for storign difference in correlation in profiles between bg pairs at each site
#generate table with all combinations of bg_1 and bg_2 for each site
diffs_expr <- data.table(expand.grid(site=unique(dt$position),bg_2=c("Wuhan-Hu-1_v2","Omicron_BA1","Omicron_BA2","Wuhan-Hu-1_v1","Alpha","Beta","Delta","Eta"),bg_1=c("Wuhan-Hu-1_v2","Omicron_BA1","Omicron_BA2","Wuhan-Hu-1_v1","Alpha","Beta","Delta","Eta")))

#remove duplicates where either bg_1 and _2 the same
diffs_expr <- diffs_expr[bg_1 != bg_2,]

#loop through and compute JSD for each site for each pair of bgs, for expr metric
diffs_expr$JSD <- as.numeric(NA) #jensen-shannon divergence, from raw expr values
diffs_expr$JSD_min3bc <- as.numeric(NA) #jensen-shannon divergence, require a minimum of 3 bcs averaged
diffs_expr$JSD_min5bc <- as.numeric(NA) #jensen-shannon divergence, require a minimum of 5 bcs averaged
for(i in 1:nrow(diffs_expr)){
  x_uncens <- dt[target==diffs_expr[i,bg_1] & position==diffs_expr[i,site],expr]
  y_uncens <- dt[target==diffs_expr[i,bg_2] & position==diffs_expr[i,site],expr]
  x_min3bc <- dt[target==diffs_expr[i,bg_1] & position==diffs_expr[i,site],expr_min3bc]
  y_min3bc <- dt[target==diffs_expr[i,bg_2] & position==diffs_expr[i,site],expr_min3bc]
  x_min5bc <- dt[target==diffs_expr[i,bg_1] & position==diffs_expr[i,site],expr_min5bc]
  y_min5bc <- dt[target==diffs_expr[i,bg_2] & position==diffs_expr[i,site],expr_min5bc]
  diffs_expr[i,JSD := JS(x_uncens,y_uncens)]
  diffs_expr[i,JSD_min3bc := JS(x_min3bc,y_min3bc)]
  diffs_expr[i,JSD_min5bc := JS(x_min3bc,y_min5bc)]
}

```

Output file with the site-pair JS divergences.
```{r save_JSD_bind_file, echo=T}
diffs_bind[,.(bg_1,bg_2,site,JSD,JSD_min3bc,JSD_min5bc)] %>%
  mutate_if(is.numeric, round, digits=6) %>%
  write.csv(file=config$JSD_v_WH1_file, row.names=F,quote=F)
```

Output file with the site-pair JS divergences.
```{r save_JSD_expr_file, echo=T}
diffs_expr[,.(bg_1,bg_2,site,JSD,JSD_min3bc,JSD_min5bc)] %>%
  mutate_if(is.numeric, round, digits=6) %>%
  write.csv(file=config$JSD_v_WH1_expr_file, row.names=F,quote=F)
```

Plotting/visualizing:

Utility function: plot scatterplot showing affinity of each of the 20 amino acids in a pair of sites

```{r scatterplot_function}
plot_scatter <- function(site, bg1, bg2, JSD=F, JSD_min3bc=T, JSD_min5bc=F,n_bc_cutoff=3,phenotype="bind"){
  x <- dt[target==bg1 & position==site,get(phenotype)]
  x_n_bc <- dt[target==bg1 & position==site,get(paste("n_bc_",phenotype,sep=""))]
  x_ref <- dt[target==bg1 & position==site & as.character(mutant)==as.character(wildtype),get(phenotype)]
  y <- dt[target==bg2 & position==site,get(phenotype)]
  y_n_bc <- dt[target==bg2 & position==site,get(paste("n_bc_",phenotype,sep=""))]
  y_ref <- dt[target==bg2 & position==site & as.character(mutant)==as.character(wildtype),get(phenotype)]
  x_min3bc <- dt[target==bg1 & position==site,get(paste(phenotype,"_min3bc",sep=""))]
  y_min3bc <- dt[target==bg2 & position==site,get(paste(phenotype,"_min3bc",sep=""))]
  x_min5bc <- dt[target==bg1 & position==site,get(paste(phenotype,"_min5bc",sep=""))]
  y_min5bc <- dt[target==bg2 & position==site,get(paste(phenotype,"_min5bc",sep=""))]
  chars <- dt[target==bg1 & position==site,mutant]
  cols <- rep("black",20); cols[which(x_n_bc < n_bc_cutoff | y_n_bc < n_bc_cutoff)] <- "orange"
  plot(x,y, xlim=if(phenotype=="bind"){c(4.5,12)}else{c(5.5,11)},ylim=if(phenotype=="bind"){c(4.5,12)}else{c(5.5,11)},pch=chars,xlab=paste(bg1,phenotype),ylab=paste(bg2,phenotype),col=cols,main=paste("site",site))
  abline(v=x_ref,lty=2,col="red")
  abline(h=y_ref,lty=2,col="red")
  if(JSD==T){
    val <- JS(x,y)
    legend("topleft",bty="n",cex=1,legend=paste("JSD:",format(val,digits=3)))
  }else if(JSD_min3bc==T){
    val <- JS(x_min3bc,y_min3bc)
    legend("topleft",bty="n",cex=1,legend=paste("JSD:",format(val,digits=3)))
  }else if(JSD_min5bc==T){
    val <- JS(x_min5bc,y_min5bc)
    legend("topleft",bty="n",cex=1,legend=paste("JSD:",format(val,digits=3)))
  }
}

```

```{r scatters_BA1_diffs, echo=T, fig.width=9, fig.height=12, fig.align="center", dpi=300,dev="png"}
par(mfrow=c(4,3))
plot_scatter(site=403,"Wuhan-Hu-1_v2","Omicron_BA1")
plot_scatter(site=406,"Wuhan-Hu-1_v2","Omicron_BA1")
plot_scatter(site=419,"Wuhan-Hu-1_v2","Omicron_BA1")
plot_scatter(site=439,"Wuhan-Hu-1_v2","Omicron_BA1")
plot_scatter(site=449,"Wuhan-Hu-1_v2","Omicron_BA1")
plot_scatter(site=455,"Wuhan-Hu-1_v2","Omicron_BA1")
plot_scatter(site=493,"Wuhan-Hu-1_v2","Omicron_BA1")
plot_scatter(site=494,"Wuhan-Hu-1_v2","Omicron_BA1")
plot_scatter(site=496,"Wuhan-Hu-1_v2","Omicron_BA1")
plot_scatter(site=498,"Wuhan-Hu-1_v2","Omicron_BA1")
plot_scatter(site=501,"Wuhan-Hu-1_v2","Omicron_BA1")
plot_scatter(site=505,"Wuhan-Hu-1_v2","Omicron_BA1")

invisible(dev.print(pdf, paste(config$epistatic_shifts_dir,"/bg-scatters_Omicron_BA1_shifts.pdf",sep=""),useDingbats=F))
```

```{r scatters_BA2_diffs, echo=T, fig.width=9, fig.height=12, fig.align="center", dpi=300,dev="png"}
par(mfrow=c(4,3))
plot_scatter(site=403,"Wuhan-Hu-1_v2","Omicron_BA2")
plot_scatter(site=406,"Wuhan-Hu-1_v2","Omicron_BA2")
plot_scatter(site=419,"Wuhan-Hu-1_v2","Omicron_BA2")
plot_scatter(site=439,"Wuhan-Hu-1_v2","Omicron_BA2")
plot_scatter(site=449,"Wuhan-Hu-1_v2","Omicron_BA2")
plot_scatter(site=455,"Wuhan-Hu-1_v2","Omicron_BA2")
plot_scatter(site=493,"Wuhan-Hu-1_v2","Omicron_BA2")
plot_scatter(site=494,"Wuhan-Hu-1_v2","Omicron_BA2")
plot_scatter(site=496,"Wuhan-Hu-1_v2","Omicron_BA2")
plot_scatter(site=498,"Wuhan-Hu-1_v2","Omicron_BA2")
plot_scatter(site=501,"Wuhan-Hu-1_v2","Omicron_BA2")
plot_scatter(site=505,"Wuhan-Hu-1_v2","Omicron_BA2")

invisible(dev.print(pdf, paste(config$epistatic_shifts_dir,"/bg-scatters_Omicron_BA2_shifts.pdf",sep=""),useDingbats=F))
```

## Line plots of JS divergence from WH1 across RBD sites

Make lineplots showing JS-D across sites for each variant compared to WH1.

Also add gray shading for sites of escape from antibodies from our large panel of antibodies we've profiled w.r.t. WH1 escape, downloaded from: https://raw.githubusercontent.com/jbloomlab/SARS2_RBD_Ab_escape_maps/main/processed_data/escape_data.csv

First, define sites of substantial antibody escape
```{r antibody_escape_sites}
dt_mAb <- data.table(read.csv(file=config$mut_antibody_escape,stringsAsFactors = F))
dt_mAb <- unique(dt_mAb[condition_type=="antibody",.(condition, condition_type, condition_subtype, site, wildtype, site_total_escape)])

dt_mAb[,site_average_escape:=mean(site_total_escape,na.rm=T),by=c("site")]

site_escape <- unique(dt_mAb[,.(wildtype, site, site_average_escape)])

#define sites for labeling as those with an average of 0.05 normalized site-wise escape across all mAbs
sig_mAb_sites <- site_escape[site_average_escape>0.15, site]


#define some epitope classes for adding highlights
label_df <- data.frame(xmin=sig_mAb_sites-0.5,
                       xmax=sig_mAb_sites+0.5)

```

```{r line_plots_JSD_v_WH1, echo=T, fig.width=12, fig.height=4, fig.align="center", dpi=300,dev="png"}
#define focal bg for others to compare to
bg <- "Wuhan-Hu-1_v2"
temp <- diffs_bind[bg_1==bg,]
temp$target <- as.character(temp$bg_2)

#define colors for each bg
group.colors <- c("Wuhan-Hu-1_v2" = cbPalette[1], "Omicron_BA1" = cbPalette[2], "Omicron_BA2" = cbPalette[8],"Wuhan-Hu-1_v1" = "black", "Alpha" = cbPalette[3], "Beta" = cbPalette[6], "Delta" = cbPalette[7], "Eta" = cbPalette[5])

#define order for plotting of bgs
temp$target <- factor(temp$target,levels=c("Alpha","Beta","Delta","Eta","Wuhan-Hu-1_v1","Omicron_BA1","Omicron_BA2"))


ggplot(data=temp, aes(x=site, y=JSD, color=target))+
  geom_rect(data=label_df, aes(x=NULL, y=NULL, color=NULL,xmin=xmin, xmax=xmax, ymin=0,ymax=1.1*max(temp$JSD,na.rm=T)), alpha=0.2)+
  geom_line(size=1)+
  scale_color_manual(values=group.colors)+
  theme_classic()+
  scale_x_continuous(expand=c(0.01,0.01),breaks=c(331,seq(335,530,by=5)))+
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.6,face="bold",size=10))+
  ylab("JS divergence versus Wuhan-Hu-1")

invisible(dev.print(pdf, paste(config$epistatic_shifts_dir,"/JSD_v_WH1.pdf",sep=""),useDingbats=F))
```

Same but require minimum 3 bc for a measurement
```{r line_plots_JSD_v_WH1_min3bc, echo=T, fig.width=12, fig.height=4, fig.align="center", dpi=300,dev="png"}
ggplot(data=temp, aes(x=site, y=JSD_min3bc, color=target))+
  geom_rect(data=label_df, aes(x=NULL, y=NULL, color=NULL,xmin=xmin, xmax=xmax, ymin=0,ymax=1.1*max(temp$JSD,na.rm=T)), alpha=0.2)+
  geom_line(size=1)+
  scale_color_manual(values=group.colors)+
  theme_classic()+
  scale_x_continuous(expand=c(0.01,0.01),breaks=c(331,seq(335,530,by=5)))+
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.6,face="bold",size=10))+
  ylab("JS divergence versus Wuhan-Hu-1")+
  geom_text_repel(aes(label=ifelse(((JSD_min3bc > 0.15)),as.character(site),'')),size=3,color="gray40")

invisible(dev.print(pdf, paste(config$epistatic_shifts_dir,"/JSD_v_WH1_min3bc.pdf",sep=""),useDingbats=F))
```

Include just Beta from the prior VOC dataset, and Beta-associated WH1 for "control"

```{r line_plots_JSD_v_WH1_min3bc_Beta-only, echo=T, fig.width=12, fig.height=4, fig.align="center", dpi=300,dev="png"}
ggplot(data=temp[bg_2 %in% c("Omicron_BA1","Omicron_BA2","Beta","Wuhan-Hu-1_v1")], aes(x=site, y=JSD_min3bc, color=target))+
  geom_rect(data=label_df, aes(x=NULL, y=NULL, color=NULL,xmin=xmin, xmax=xmax, ymin=0,ymax=1.1*max(temp$JSD,na.rm=T)), alpha=0.2)+
  geom_line(size=1)+
  scale_color_manual(values=group.colors)+
  theme_classic()+
  scale_x_continuous(expand=c(0.01,0.01),breaks=c(331,seq(335,530,by=5)))+
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.6,face="bold",size=10))+
  ylab("JS divergence versus Wuhan-Hu-1")+
  geom_text_repel(aes(label=ifelse(((JSD_min3bc > 0.15)),as.character(site),'')),size=3,color="gray40")

invisible(dev.print(pdf, paste(config$epistatic_shifts_dir,"/JSD_v_WH1_min3bc_Omi-and-Beta-only.pdf",sep=""),useDingbats=F))
```

Scatterplots to more clearly show the additional shifts in each Omicron variant versus Beta?

```{r scatterplots_epistatic-shift_Beta-BA1-BA2, echo=T, fig.width=12, fig.height=4, fig.align="center", dpi=300,dev="png"}
temp2 <- dcast(temp[bg_2 %in% c("Omicron_BA1","Omicron_BA2","Beta")], site ~ bg_2, value.var="JSD_min3bc")

p1 <- ggplot(data=temp2, aes(x=Omicron_BA1, y=Omicron_BA2))+
  geom_point()+
  geom_text_repel(aes(label=ifelse((Omicron_BA2 > 0.1 | Omicron_BA1 > 0.1),as.character(site),'')),size=3)+
  theme_classic()+
  xlab("epistatic shift versus Wuhan-Hu-1, Omicron BA.1")+
  ylab("epistatic shift versus Wuhan-Hu-1, Omicron BA.2")

p2 <- ggplot(data=temp2, aes(x=Omicron_BA1, y=Beta))+
  geom_point()+
  geom_text_repel(aes(label=ifelse((Omicron_BA1 > 0.1 | Beta > 0.1),as.character(site),'')),size=3)+
  theme_classic()+
  xlab("epistatic shift versus Wuhan-Hu-1, Omicron BA.1")+
  ylab("epistatic shift versus Wuhan-Hu-1, Beta")

p3 <- ggplot(data=temp2, aes(x=Omicron_BA2, y=Beta))+
  geom_point()+
  geom_text_repel(aes(label=ifelse((Omicron_BA2 > 0.1 | Beta > 0.1),as.character(site),'')),size=3)+
  theme_classic()+
  xlab("epistatic shift versus Wuhan-Hu-1, Omicron BA.2")+
  ylab("epistatic shift versus Wuhan-Hu-1, Beta")

grid.arrange(p1,p2,p3,nrow=1)

invisible(dev.print(pdf, paste(config$epistatic_shifts_dir,"/JSD_v_WH1_min3bc_Omi-versus-Beta-scatters.pdf",sep=""),useDingbats=F))
```

Repeat for expression measurements

```{r line_plots_JSD_v_WH1_expr, echo=T, fig.width=12, fig.height=4, fig.align="center", dpi=300,dev="png"}
#define focal bg for others to compare to
bg <- "Wuhan-Hu-1_v2"
temp <- diffs_expr[bg_1==bg,]
temp$target <- as.character(temp$bg_2)

#define order for plotting of bgs
temp$target <- factor(temp$target,levels=c("Alpha","Beta","Delta","Eta","Wuhan-Hu-1_v1","Omicron_BA1","Omicron_BA2"))


ggplot(data=temp, aes(x=site, y=JSD, color=target))+
  geom_rect(data=label_df, aes(x=NULL, y=NULL, color=NULL,xmin=xmin, xmax=xmax, ymin=0,ymax=1.1*max(temp$JSD,na.rm=T)), alpha=0.2)+
  geom_line(size=1)+
  scale_color_manual(values=group.colors)+
  theme_classic()+
  scale_x_continuous(expand=c(0.01,0.01),breaks=c(331,seq(335,530,by=5)))+
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.6,face="bold",size=10))+
  ylab("JS divergence versus Wuhan-Hu-1")

invisible(dev.print(pdf, paste(config$epistatic_shifts_dir,"/JSD_v_WH1_expr.pdf",sep=""),useDingbats=F))
```

Same but require minimum 3 bc for a measurement
```{r line_plots_JSD_v_WH1_min3bc_expr, echo=T, fig.width=12, fig.height=4, fig.align="center", dpi=300,dev="png"}
ggplot(data=temp, aes(x=site, y=JSD_min3bc, color=target))+
  geom_rect(data=label_df, aes(x=NULL, y=NULL, color=NULL,xmin=xmin, xmax=xmax, ymin=0,ymax=1.1*max(temp$JSD,na.rm=T)), alpha=0.2)+
  geom_line(size=1)+
  scale_color_manual(values=group.colors)+
  theme_classic()+
  scale_x_continuous(expand=c(0.01,0.01),breaks=c(331,seq(335,530,by=5)))+
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.6,face="bold",size=10))+
  ylab("JS divergence versus Wuhan-Hu-1, expression DMS")+
  geom_text_repel(aes(label=ifelse(((JSD_min3bc > 0.15)),as.character(site),'')),size=3,color="gray40")

invisible(dev.print(pdf, paste(config$epistatic_shifts_dir,"/JSD_v_WH1_min3bc_expr.pdf",sep=""),useDingbats=F))
```

Include just Beta from the prior VOC dataset, and Beta-associated Wuhan-Hu-1 measurements for "control"

```{r line_plots_JSD_v_WH1_expr_min3bc_Beta-only, echo=T, fig.width=12, fig.height=4, fig.align="center", dpi=300,dev="png"}
ggplot(data=temp[bg_2 %in% c("Omicron_BA1","Omicron_BA2","Beta","Wuhan-Hu-1_v1")], aes(x=site, y=JSD_min3bc, color=target))+
  geom_rect(data=label_df, aes(x=NULL, y=NULL, color=NULL,xmin=xmin, xmax=xmax, ymin=0,ymax=1.1*max(temp$JSD,na.rm=T)), alpha=0.2)+
  geom_line(size=1)+
  scale_color_manual(values=group.colors)+
  theme_classic()+
  scale_x_continuous(expand=c(0.01,0.01),breaks=c(331,seq(335,530,by=5)))+
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.6,face="bold",size=10))+
  ylab("JS divergence versus Wuhan-Hu-1")+
  geom_text_repel(aes(label=ifelse(((JSD_min3bc > 0.1)),as.character(site),'')),size=3,color="gray40")

invisible(dev.print(pdf, paste(config$epistatic_shifts_dir,"/JSD_v_WH1_min3bc_expr_Omi-and-Beta-only.pdf",sep=""),useDingbats=F))
```

```{r scatterplots_epistatic-shift_Beta-BA1-BA2_expr, echo=T, fig.width=16, fig.height=4, fig.align="center", dpi=300,dev="png"}
temp[bg_2=="Wuhan-Hu-1_v1",bg_2:="Wuhan.Hu.1_v1"]
temp2 <- dcast(temp[bg_2 %in% c("Omicron_BA1","Omicron_BA2","Beta","Wuhan.Hu.1_v1")], site ~ bg_2, value.var="JSD_min3bc")

p1 <- ggplot(data=temp2, aes(x=Omicron_BA1, y=Omicron_BA2))+
  geom_point()+
  geom_text_repel(aes(label=ifelse((Omicron_BA2 > 0.1 | Omicron_BA1 > 0.1),as.character(site),'')),size=3)+
  theme_classic()+
  scale_x_continuous(limits = c(0, 0.31))+
  scale_y_continuous(limits = c(0, 0.31))

p2 <- ggplot(data=temp2, aes(x=Omicron_BA1, y=Beta))+
  geom_point()+
  geom_text_repel(aes(label=ifelse((Omicron_BA1 > 0.1 | Beta > 0.1),as.character(site),'')),size=3)+
  theme_classic()+
  scale_x_continuous(limits = c(0, 0.31))+
  scale_y_continuous(limits = c(0, 0.31))

p3 <- ggplot(data=temp2, aes(x=Omicron_BA2, y=Beta))+
  geom_point()+
  geom_text_repel(aes(label=ifelse((Omicron_BA2 > 0.1 | Beta > 0.1),as.character(site),'')),size=3)+
  theme_classic()+
  scale_x_continuous(limits = c(0, 0.31))+
  scale_y_continuous(limits = c(0, 0.31))

p4 <- ggplot(data=temp2, aes(x=Omicron_BA1, y=Wuhan.Hu.1_v1))+
  geom_point()+
  geom_text_repel(aes(label=ifelse((Wuhan.Hu.1_v1 > 0.1 | Beta > 0.1),as.character(site),'')),size=3)+
  theme_classic()+
  scale_x_continuous(limits = c(0, 0.31))+
  scale_y_continuous(limits = c(0, 0.31))

grid.arrange(p1,p2,p3,p4,nrow=1)

invisible(dev.print(pdf, paste(config$epistatic_shifts_dir,"/JSD_v_WH1_min3bc_expr_Omi-versus-Beta-scatters.pdf",sep=""),useDingbats=F))
```

## Map divergence to pdb structure

First, bind
```{r map_JSD_to_pdb}
pdb_wh1 <- read.pdb(file=config$pdb_6m0j)

#iterate through backgrounds, output a pdb comparing its divergence to WH1 (using min3bc)
for(s in c("Omicron_BA1","Omicron_BA2")){
  b <- rep(0, length(pdb_wh1$atom$b))
  for(i in 1:nrow(pdb_wh1$atom)){
    if(pdb_wh1$atom$chain[i]=="E"){
      res <- pdb_wh1$atom$resno[i]
      JSD <- diffs_bind[bg_1=="Wuhan-Hu-1_v2" & bg_2==s & site==res, JSD_min3bc]
      if(length(JSD)>0){
        b[i] <- JSD
      }
    }
  }
  write.pdb(pdb=pdb_wh1, file=paste(config$epistatic_shifts_dir,"/pdbs/",s,"_v_WH1_JSD-min3bc.pdb",sep=""), b=b)
}

```
repeat for expression measures
```{r map_JSD_to_pdb_expr}
pdb_wh1 <- read.pdb(file=config$pdb_6m0j)

#iterate through backgrounds, output a pdb comparing its divergence to WH1 (using min3bc)
for(s in c("Omicron_BA1","Omicron_BA2")){
  b <- rep(0, length(pdb_wh1$atom$b))
  for(i in 1:nrow(pdb_wh1$atom)){
    if(pdb_wh1$atom$chain[i]=="E"){
      res <- pdb_wh1$atom$resno[i]
      JSD <- diffs_expr[bg_1=="Wuhan-Hu-1_v2" & bg_2==s & site==res, JSD_min3bc]
      if(length(JSD)>0){
        b[i] <- JSD
      }
    }
  }
  write.pdb(pdb=pdb_wh1, file=paste(config$epistatic_shifts_dir,"/pdbs/",s,"_v_WH1_JSD-min3bc_expr.pdb",sep=""), b=b)
}
```

```{r look more at 493, fig.width=8, fig.height=4, fig.align="center", dpi=300,dev="png"}
par(mfrow=c(1,2))
temp <- dt[position==493 & target %in% c("Omicron_BA1","Wuhan-Hu-1_v2"),]

temp[,delta_bind_Q493ref := as.numeric(NA)]
for(i in 1:nrow(temp)){
  temp$delta_bind_Q493ref[i] <- temp$bind[i] - temp[target==temp$target[i] & mutant=="Q",bind] 
}

plot(temp[target=="Wuhan-Hu-1_v2",delta_bind_Q493ref],temp[target=="Omicron_BA1",delta_bind_Q493ref],pch=temp[target=="Wuhan-Hu-1_v2",mutant],xlab="Q493x mutation, WH1",ylab="Q493x mutation, BA1",xlim=c(-2.5,1.7),ylim=c(-2.5,1.7))
abline(a=0,b=1)
abline(v=0,lty=2)
abline(h=0,lty=2)

temp2 <- dt[position==493 & target %in% c("Omicron_BA2","Wuhan-Hu-1_v2"),]

temp2[,delta_bind_Q493ref := as.numeric(NA)]
for(i in 1:nrow(temp2)){
  temp2$delta_bind_Q493ref[i] <- temp2$bind[i] - temp2[target==temp2$target[i] & mutant=="Q",bind] 
}

plot(temp2[target=="Wuhan-Hu-1_v2",delta_bind_Q493ref],temp2[target=="Omicron_BA2",delta_bind_Q493ref],pch=temp2[target=="Wuhan-Hu-1_v2",mutant],xlab="Q493x mutation, WH1",ylab="Q493x mutation, BA1",xlim=c(-2.5,1.7),ylim=c(-2.5,1.7))
abline(a=0,b=1)
abline(v=0,lty=2)
abline(h=0,lty=2)


invisible(dev.print(pdf, paste(config$epistatic_shifts_dir,"/scatter_Q493x_mut.pdf",sep=""),useDingbats=F))
```

For each position that varies between WH1 and BA.1 and BA.2, make little plots that illustrate epistasis of the substitution itself to show entrenchment and anti-entrenchment

```{r entrenchment_Omicron_BA1, fig.width=7.5, fig.height=12.5, fig.align="center", dpi=300,dev="png"}
entrench_BA1 <- data.table(site=NA,WH1_wt=NA,BA1_wt=NA,WH1_wt_bind=NA,WH1_mut_bind=NA,BA1_revert_bind=NA,BA1_wt_bind=NA)

for(pos in unique(dt$position)){
  WH1_wt <- dt[target=="Wuhan-Hu-1_v2" & wildtype==mutant & position==pos,wildtype]
  BA1_wt <- dt[target=="Omicron_BA1" & wildtype==mutant & position==pos,wildtype]
  if(WH1_wt != BA1_wt){
    entrench_BA1 <- rbind(entrench_BA1, 
                          list(site=pos,
                               WH1_wt=WH1_wt,
                               BA1_wt=BA1_wt,
                               WH1_wt_bind=dt[target=="Wuhan-Hu-1_v2" & position==pos & mutant==WH1_wt,bind],
                               WH1_mut_bind=dt[target=="Wuhan-Hu-1_v2" & position==pos & mutant==BA1_wt,bind],
                               BA1_revert_bind=dt[target=="Omicron_BA1" & position==pos & mutant==WH1_wt,bind],
                               BA1_wt_bind=dt[target=="Omicron_BA1" & position==pos & mutant==BA1_wt,bind]
                               )
                          )
  }
}
entrench_BA1 <- entrench_BA1[-1,]

par(mfrow=c(5,3))
for(i in 1:nrow(entrench_BA1)){
  plot(1:4, 
       entrench_BA1[i,4:7],
       main=paste(entrench_BA1[i,c(2,1,3)],collapse=""),
       pch=16,cex=2,
       col=c(cbbPalette[1],cbbPalette[1],cbbPalette[2],cbbPalette[2]),
       xlim=c(0.75, 4.25), ylim=c(6,10.5),
       ylab="ACE2 affinity (-log10Kd)",
       xlab="amino acid",
       xaxt="n")
  axis(1, at=1:4, labels=c("WH1", "WH1+\nmut","BA.1+\nrevert","BA.1"))
  points(1:2, entrench_BA1[i,4:5],type="l",lwd=1.5)
  points(3:4, entrench_BA1[i,6:7],type="l",lwd=1.5,col=cbbPalette[2])
}

invisible(dev.print(pdf, paste(config$epistatic_shifts_dir,"/entrenchment_diagrams_BA1.pdf",sep=""),useDingbats=F))

```



```{r entrenchment_Omicron_BA2, fig.width=7.5, fig.height=15, fig.align="center", dpi=300,dev="png"}
entrench_BA2 <- data.table(site=NA,WH1_wt=NA,BA2_wt=NA,WH1_wt_bind=NA,WH1_mut_bind=NA,BA2_revert_bind=NA,BA2_wt_bind=NA)

for(pos in unique(dt$position)){
  WH1_wt <- dt[target=="Wuhan-Hu-1_v2" & wildtype==mutant & position==pos,wildtype]
  BA2_wt <- dt[target=="Omicron_BA2" & wildtype==mutant & position==pos,wildtype]
  if(WH1_wt != BA2_wt){
    entrench_BA2 <- rbind(entrench_BA2, 
                          list(site=pos,
                               WH1_wt=WH1_wt,
                               BA2_wt=BA2_wt,
                               WH1_wt_bind=dt[target=="Wuhan-Hu-1_v2" & position==pos & mutant==WH1_wt,bind],
                               WH1_mut_bind=dt[target=="Wuhan-Hu-1_v2" & position==pos & mutant==BA2_wt,bind],
                               BA2_revert_bind=dt[target=="Omicron_BA2" & position==pos & mutant==WH1_wt,bind],
                               BA2_wt_bind=dt[target=="Omicron_BA2" & position==pos & mutant==BA2_wt,bind]
                               )
                          )
  }
}
entrench_BA2 <- entrench_BA2[-1,]

par(mfrow=c(6,3))
for(i in 1:nrow(entrench_BA2)){
  plot(1:4, 
       entrench_BA2[i,4:7],
       main=paste(entrench_BA2[i,c(2,1,3)],collapse=""),
       pch=16,cex=2,
       col=c(cbbPalette[1],cbbPalette[1],cbbPalette[8],cbbPalette[8]),
       xlim=c(0.75, 4.25), ylim=c(6,10.5),
       ylab="ACE2 affinity (-log10Kd)",
       xlab="",
       xaxt="n")
  axis(1, at=1:4, labels=c("WH1", "WH1+\nmut","BA.2+\nrevert","BA.2"))
  points(1:2, entrench_BA2[i,4:5],type="l",lwd=1.5)
  points(3:4, entrench_BA2[i,6:7],type="l",lwd=1.5,col=cbbPalette[8])
}

invisible(dev.print(pdf, paste(config$epistatic_shifts_dir,"/entrenchment_diagrams_BA2.pdf",sep=""),useDingbats=F))

```






